---
title: "cascadeR"
author: "Apratim Mitra"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  `cascadeR` is an interactive & modular Shiny app that simplifies and transforms
  complex single-cell RNA-Seq and spatial transcriptomics data using numerous
  interactive features. Designed for both computational and experimental
  biologists, `cascadeR` supports both R and python ecosystems, making data exploration
  and accessible, while also providing a platform to manage multiple datasets either
  locally or on a server to share with collaborators (package version:
  `r packageVersion("cascadeR")`).
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{cascadeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Install cascadeR

Install cascadeR from Bioconductor using `BiocManager::install`.

```{r pkg_install, eval=FALSE}
# first check to see if BiocManager is available
if(!requireNamespace('BiocManager', quietly=TRUE)){
  install.packages('BiocManager')
}

BiocManager::install('cascadeR')
```

## Get example data

Now we load a small dataset with 230 cells. It contains
two dimension reductions: *pca* and *tsne*.

```{r load}
library(Seurat)

data("pbmc_small", package="SeuratObject")

pbmc_small
```

## Inspect the metadata

Next, let's take a look at the metadata of this object.

```{r metadata}
head(pbmc_small)
```

There are two columns with standard quality control metrics:

- *nCount_RNA*
- *nFeature_RNA*

In addition, there are two columns with cell clusters calculated
at two resolutions. Higher resolutions correspond to more/smaller
cell clusters.

- *RNA_snn_res.0.8*: resolution = 0.8, 2 clusters
- *RNA_snn_res.1*: resolution = 1, 3 clusters

We visualize the resolution = `1.0` clustering using `Seurat::DimPlot`:

```{r dimplot}
Seurat::DimPlot(pbmc_small, group.by='RNA_snn_res.1')
```

## Marker detection

Next, we calculate gene markers to characterize individual
clusters for one of the resolutions using the `Seurat::FindAllMarkers`
function. Setting `Idents()` here to `RNA_snn_res.0.8` sets the
cell groups to clusters at resolution = `0.8`.

```{r markers}
Idents(pbmc_small) <- pbmc_small$RNA_snn_res.0.8
df <- FindAllMarkers(pbmc_small)
```

Now we have marker genes for each of the clusters at this resolution.

```{r head_markers}
head(df)
```

We often like to compare different resolutions, so let's do the
same for resolution = `1.0`.

```{r markers2, warning=FALSE}
Idents(pbmc_small) <- pbmc_small$RNA_snn_res.1
df2 <- FindAllMarkers(pbmc_small)
```

You can combine multiple such marker gene tables for use with `cascadeR`
to compare, e.g. different resolutions, "assays" or arbitrary "groups".
To do this, we concatenate our two marker gene sets and add a
"resolution" column.

```{r join_df}
df$resolution <- '0.8'
df2$resolution <- '1.0'

marker_df <- rbind(df, df2)
```


## Save results

Next, we save our results.

```{r save}
saveRDS(pbmc_small, 'seurat_pbmc_small.Rds', compress=FALSE)
write.table(marker_df, 'allmarkers_combined.tsv', sep='\t', row.names=FALSE, quote=FALSE)
```

## Data Organization

Next, we need to organize your data in a directory structure that `cascadeR`
can easily navigate, e.g. in a folder `cascader/data` in your home directory.
The folder should look something like this:

```
~/cascader/data/
  └─ my-project
     └─ pbmc_small
        ├─ clustered.Rds
        └─ allmarkers.tsv
```

Two things to note here:

- *Projects* are subfolders in the data area, while *analyses* are
  subfolders within projects. For instance, in the above example,
  `my-project` is the project name and `pbmc_small` is the analysis name.
  This 2-level hierarchy allows us to have multiple analysis using, e.g
  different analysis parameters, within a single project.

- `cascadeR` looks for exact filenames, e.g. `clustered.Rds`, when figuring
  out what data to read. Here is a list currently supported:

  - Single-cell object: `clustered.Rds` (Seurat) or `clustered.h5ad` (anndata)
  - `allmarkers.tsv`: Marker genes for a cluster (compared to everything else).
    Output from Seurat's `FindMarkers` or `FindAllMarkers` or scanpy's
    `sc.tl.rank_genes_groups` are supported.
  - `consmarkers.tsv`: Conserved marker genes across samples, output from
    Seurat's `FindConservedMarkers`.
  - `demarkers.tsv`: Differential expression analysis, e.g. from comparing
    cell clusters between samples. Output from Seurat's `FindMarkers` or
    output from pseudo-bulk DE analysis from `DESeq2`, `edgeR`, `limma`
    supported.

  As mentioned above, each of the marker gene tables supports special columns,
  "resolution", "assay", "group" to combine multiple sets of results
  for joint viewing.

We provide a helper function to set up a dataset while following the above
specifications. Let's use this now for the analysis we just performed.

First we create a data directory for `cascadeR`.

```{r create_dir, eval=FALSE}
dir.create('~/cascader/data', recursive=TRUE)
```

Next, we run `add_cascade_analysis` with the RDS file and marker table path
specified. By default, this function prints out the commands to be run -
to actually run them, we use `execute = TRUE`.

```{r add_analysis, eval=FALSE}
add_cascade_analysis(
  obj_path='seurat_pbmc_small.Rds',
  data_dir='~/cascader/data',
  project='my-project',
  analysis='pbmc_small',
  cluster_markers='allmarkers_combined.tsv',
  execute=TRUE
)
```

## First Run

Load the `cascadeR` package.

```{r first_load, eval=FALSE}
library(cascadeR)
```

# sessionInfo

```{r}
sessionInfo()
```
